---
title: "Crime in Chicago: A Data Analysis Summary using R"
author: 'Student ID: 201081646'
output:
  pdf_document:
    fig_caption: yes
    fig_crop: yes
    fig_height: 3.3
    fig_width: 5.3
    keep_tex: yes
    number_sections: yes
    toc_depth: 2
  html_document:
    fig_caption: yes
    theme: journal
    toc: yes
    toc_depth: 2
header-includes:
 \usepackage{float}
 \floatplacement{figure}{H}
---
```{r, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

# Introduction

This report is the second assessment of the **MATH5741M Statistical Theory and Methods** module. Its aim is to analyse a vehicle accident dataset from the Department for Transport and performe the following tasks:

- Create a boxplot 

- Statistical hypothesis test

- confidence interval

# Data preparation

First, we activate the libraries we will need to set up the project.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Activate libraries
library(ggplot2)
library(lubridate)
library(zoo)
library(dplyr)
library(knitr)
```

Second, we load the data into the `R` environment.

```{r eval=TRUE, echo=TRUE}
# Read csv in R
xx=read.csv("http://www1.maths.leeds.ac.uk/~charles/math5741/DfTaccidents.csv",header=T)
```

The next step is to drop all those columns we do not need to answer our research questions.

```{r, echo=TRUE}
# Drop all variables we are not interested in
xx <- xx[, -c(1:2, 4:11)]
```

Then, we clean the dataset of missing values.

```{r, echo=TRUE}
# Remove NAs
xx <- xx[complete.cases(xx),]
```

Finally, we show the the dataset ready for exploration.

```{r, echo=TRUE}
# Show first 5 records
head(xx)
```
```{r, echo=TRUE}
# Show first 5 records
names(xx)
```
# Data exploration

## Boxplot

```{r, echo=TRUE}

# list urbans or rural variable
xx$Urban_or_Rural_Area[xx$Urban_or_Rural_Area == "1"] <- "Urban"
xx$Urban_or_Rural_Area[xx$Urban_or_Rural_Area == "2"] <- "Rural"
xx$Urban_or_Rural_Area[xx$Urban_or_Rural_Area == "3"] <- "Unallocated"

# Remove Unallocated recods


# Draw a boxplot
ggplot(xx, aes(x=Urban_or_Rural_Area, y=Number_of_Vehicles))+geom_boxplot()

ggplot(xx, aes(x=Urban_or_Rural_Area, y=Number_of_Vehicles))+geom_boxplot()+scale_y_log10()
```

## statistical hypothesis


## confidence interval
