---
title: "Road Traffic Accidents Data Analysis using R"
author: 'Student ID: 201081646'
header-includes: \usepackage{float} \floatplacement{figure}{H}
output:
  pdf_document:
    fig_caption: yes
    fig_crop: yes
    fig_height: 3.3
    fig_width: 5.3
    keep_tex: yes
    number_sections: yes
    toc_depth: 2
  html_document:
    fig_caption: yes
    theme: journal
    toc: yes
    toc_depth: 2
  github_document:
    toc_depth: 2
  word_document:
    toc: yes
    toc_depth: '2'
bibliography: Stats.bib
---

```{r, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

# Introduction

This report is the second assessment of the **MATH5741M Statistical Theory and Methods** module. Its aim is to answer three questions through inferential statistical analysis regarding a road traffic accidents dataset from 2005 collected by the UK Department for Transport.

All the analysis has been done using **R** (programming language) and is code reproducible. To see the complete **R** coding process visit https://github.com/eugenividal/Road-Traffic-Accidents-Data-Analysis

```{r, message=FALSE, warning=FALSE, include=FALSE}
# Activate libraries
library(tidyverse)
library(cowplot)
```

```{r eval=TRUE, include=FALSE}
# Read csv in R
xx=read.csv("DfTaccidents.csv", header=T)
```

```{r, message=FALSE, warning=FALSE, include=FALSE}
# Have a look at the variables we got
names(xx)
# Drop all those variables we do not need
xx <- xx[, c(8, 10, 11, 30)]
# Create a variable count with value 1 (for future aggregates)
xx$count <- 1 
```

```{r, include=FALSE}
# Rename labels of Urban_or_Rural_Area
xx$Urban_or_Rural_Area[xx$Urban_or_Rural_Area == "1"] <- "Urban"
xx$Urban_or_Rural_Area[xx$Urban_or_Rural_Area == "2"] <- "Rural"
xx$Urban_or_Rural_Area[xx$Urban_or_Rural_Area == "3"] <- "Unallocated"

# Rename labels of day of the week
xx$Day_of_Week[xx$Day_of_Week == "1"] <- "Sunday"
xx$Day_of_Week[xx$Day_of_Week == "2"] <- "Monday"
xx$Day_of_Week[xx$Day_of_Week == "3"] <- "Tuesday"
xx$Day_of_Week[xx$Day_of_Week == "4"] <- "Wednesday"
xx$Day_of_Week[xx$Day_of_Week == "5"] <- "Thursday"
xx$Day_of_Week[xx$Day_of_Week == "6"] <- "Friday"
xx$Day_of_Week[xx$Day_of_Week == "7"] <- "Saturday"
```

```{r, message=FALSE, warning=FALSE, include=FALSE}
# Data ready for analysis - show first 6 rows
head(xx)
```

# Results

## Question 1

<!--(1)Draw a boxplot to compare the number of vehicles involved in an urban area, with the number involved in a rural area. (2) Explain why a transformation of the data may (or may not) be appropriate. Using your transformation (or not) (3) carry out a suitable test to investigate whether the average number of vehicles in an accident differs in urban and rural areas.-->

In this question, first, we are asked to draw a boxplot to compare the number of vehicles involved in urban areas with the number invlolved in rural areas. Secondly, we have to carry out a test to investigate whether the average number of vehicles in an accident differs per type of area.

<!--http://www.jbstatistics.com/pooled-variance-t-tests-and-confidence-intervals-an-example/-->

To plot the boxplot, we first remove the unallocated values. 

```{r, fig.align='center', message=FALSE, warning=FALSE, include=FALSE, out.extra=''}
# Remove Unallocated recods
xx_a = xx[!xx$Urban_or_Rural_Area == "Unallocated",]
```

```{r, echo=FALSE, fig.align='center', message=FALSE, warning=FALSE, out.extra=''}
# Draw a boxplot
p <- ggplot(xx_a, aes(x=Urban_or_Rural_Area, y=Number_of_Vehicles))+geom_boxplot()
p
```

IN the boxplot is difficult to appreciate what is going on. So, we visualise the data to analyse if it is normal (histogram 1). Because it is very skewed to the right, we take the log10 and see that it substantially improves (histogram 2). <!-- should I show table of values or plot a QQ plot instead?-->  

```{r}
# show table of values
table(xx_a$Number_of_Vehicles)
```

```{r, include=FALSE}
# Summaries
summary(xx_a$Number_of_Vehicles)
summary(log10(xx_a$Number_of_Vehicles))
xx_a$log10 <- log10(xx_a$Number_of_Vehicles)
```

```{r, fig, fig.cap="Histograms", echo=FALSE, message=FALSE, warning=FALSE, out.extra='',fig.align='center'}
# Histograms
p2 <- ggplot(xx_a, aes(x=Number_of_Vehicles)) +
    geom_histogram(position="identity", colour="grey40", alpha=0.2, bins = 10) +
    facet_grid(. ~ Urban_or_Rural_Area) +  xlab("Type of area") + ylab("Number of vehicles")+  theme_minimal(base_size = 8)
p3<-p2 + scale_y_continuous(trans = "log10")
p3<-p3 +  xlab("Type of area") + ylab("Number of vehicles log10")
p4<-plot_grid(p2, p3, labels = c("H1","H2"), align = "h", label_size = 10, label_x = 0, label_y = 0, hjust = -0.5, vjust = -0.5)
p4  
```

```{r, fig2, fig.cap="Number of vehicles involved grouped by type of area", echo=FALSE, message=FALSE, warning=FALSE, out.extra='',fig.align='center'}
# Boxplot with data transformed
p6 <- ggplot(xx_a, aes(x=Urban_or_Rural_Area, y=Number_of_Vehicles))+geom_boxplot()+scale_y_log10()
p6 +  xlab("Type of area") + ylab("Number of vehicles -log10") +  theme_minimal(base_size = 8)
```

From the boxplot we cannot be sure that the populations differ.

So, we carry out a test to investigate wheter the average number of vehicles in an accident differs per type of area. 

http://www.jbstatistics.com/inference-for-two-means-introduction/

Determination of hypotheses:

$$H_{0}: \mu_{urban} = \mu_{rural}\;\;\;vs.\;\;\;H_{1}: \mu_{urban} \neq \mu_{rural}$$

The summary statistics are this summary()

```{r, message=FALSE, warning=FALSE, include=FALSE}
# Check the variance of the two groups
xx_u <- xx_a%>%
  filter(Urban_or_Rural_Area=="Urban")
xx_r <- xx_a%>%
  filter(Urban_or_Rural_Area=="Rural")
n1= 72267
n2= 126378
xbaru= mean(xx_u$log10)
xbaru
xbarr= mean(xx_r$log10)
xbarr
su= sd(xx_u$log10)
su
sr= sd(xx_r$log10)
sr
```

It seems reasonable to assume \(\sigma_{1}^{2}\) = \(\sigma_{2}^{2}\) and so we estimate the common variance with a  two-sided test for two-sample poblems  (*pooled estimate*). <!--As a rough rule-of thumb, you can assume equal variances when the ratio of max/min is less than 3 (or less than 4 for small samples)-->

The test statistics is then:

We do not reject the null hypotesis and conclude that \(\mu_{urban}\) = \(\mu_{rural}\).

## Question 2

<!--Using a suitable statistical hypothesis test, investigate whether the frequency of accidents varies by day of the week. Repeat this test using only week-days (excuding Saturday and Sunday).-->

In this question, first, we have to investigate whether the frequency of accidents varies by day of the week using the suitable statistical hypothesis test.

For this, we apply a **chi-squared test**. Check example p.91.

http://www.jbstatistics.com/chi-square-tests-for-one-way-tables/

```{r, include=FALSE}
xx_group_alldays<- xx%>%
  select(Day_of_Week, Date, count)%>%
  group_by(Date, Day_of_Week)%>%
  summarise(n = sum(count))
chisq.test(xx_group_alldays$n)
```

Next, we are required to do the same test using only week-days (excluding Saturday and Sunday).

```{r, include=FALSE}
# Select data using only week-days
xx_group_weekdays <- xx_group_alldays[!xx_group_alldays$Day_of_Week == "Saturday",]
xx_group_weekdays <- xx_group_weekdays[!xx_group_weekdays$Day_of_Week == "Sunday",]
# hypothesis per each day using only week-days
#shapiro.test(xx_group_weekdays$n) #Since the p-value is greater than 0.05 we accept that the data come from a normal distribution
chisq.test(xx_group_weekdays$n)
```

## Question 3

<!--Compute a 95% confidence interval for the expected (mean) number of accidents which occur on a Monday. State your assumptions in computing this interval, and verify whether they are valid.-->

Finally, in question 3, we are asked to compute a 95% confidence interval for the expected (mean) number of accidents which occur on a Monday.

Look at example in p.62 to answer this:

- The data is clearly not normal. The distribution is discrete and very skewed to the right. The table of values is:

- The sample is huge (*n*= 198,735) so the central limit theorem stats that the sample mean in normallly distributed (even when the data are not).

- However, we need to think - for which population are we estimating a mean for?.

Is this a simple t-distribution confidence interval? WE could say that because the sample is that bing could be just a normal confidence interval?

```{r, echo=TRUE}
# Select data using only week-days
xx_mon<- xx%>%
  select(Day_of_Week, Date, count)%>%
  filter(Day_of_Week=="Monday")%>%
  group_by(Date)%>%
  summarise(n = sum(count))
```

```{r}
#t.test(Mondays, conf.level = 0.95)$conf.int
t.test(xx_mon$n)$conf.int
```

# Bibliography

The resources used to carry out this project were:

- Balka, J. 2013. JBStatistics: Making Statistics Make Sense. Available from: http://www.jbstatistics.com.
- Lane, D.M. 2018. Online Statistics Education: An Interactive Multimedia Course of Study. Available from: http://onlinestatbook.com/.
- Taylor, C. 2017. MATH5741M: Statistical Theory and Methods. Outline Lecture Notes.
- Yau, C. 2018. R tutorial: an R introduction to statistics. Available from: http://www.r-tutor.com