---
title: "Road Traffic Accidents Data Analysis using R"
author: 'Student ID: 201081646'
output: github_document
#  output:
#   pdf_document:
#     fig_caption: yes
#     fig_crop: yes
#     fig_height: 3.3
#     fig_width: 5.3
#     keep_tex: yes
#     number_sections: yes
#     toc_depth: 2
#   html_document:
#     fig_caption: yes
#     theme: journal
#     toc: yes
#     toc_depth: 2
# header-includes:
#  \usepackage{float}
#  \floatplacement{figure}{H}
---

```{r, include=FALSE}
knitr::opts_chunk$set(fig.pos = 'H')
```

# Introduction

This report is the second assessment of the **MATH5741M Statistical Theory and Methods** module. Its aim is to analyse a road traffic accidents dataset collected by the UK Department for Transport (DfT) in 2005 trough different statistical methods such as: Boxplot visualisations, Statistical hypothesis testing, and Computing Confidence interval.

It has been done using **R** (programming language) and it is code reproducible. To see the whole code written for its performance visit https://github.com/eugenividal/Road-Traffic-Accidents-Data-Analysis

# Data preparation

First, we activate the libraries we will need to set up the project.

```{r, echo=TRUE, message=FALSE, warning=FALSE}
# Activate libraries
library(ggplot2)
```

Second, we load the data into the **R** environment.

```{r eval=TRUE, echo=TRUE}
# Read csv in R
#xx=read.csv("http://www1.maths.leeds.ac.uk/~charles/math5741/DfTaccidents.csv",header=T)
xx=read.csv("DfTaccidents.csv", header=T)
```

The next step is to drop all those columns we will not need to perform our analysis. We only need: number of vehicles, type of area and day of the week.

```{r, message=FALSE, warning=FALSE, include=FALSE}
# Have a look at the variables we got
names(xx)
# Drop all those we do not need
xx <- xx[, c(8, 11, 30)]
```

Finally, we show the the dataset ready for exploration.

```{r, echo=TRUE}
# Show first 6 rows
head(xx)
```

# Data exploration

## Boxplot

**Draw a boxplot** to compare the number of vehicles involved in an urban area, with the number involved in a rural area. **Explain why a transformation** of the data may (or may not) be appropriate. Using your transformation (or not) **carry out a suitable test** to investigate whether the average number of vehicles in an accident differs in urban and rural areas.

The following boxplot compare the number of vehicles involved in an urban area with the number involved in a rural area. The data has been transformed in order to..., otherwise we cannot apreciate well the...

```{r, fig, fig.cap="Number of vehicles involved grouped by type of area", echo=FALSE, message=FALSE, warning=FALSE, out.extra='',fig.align='center'}

# List urbans or rural variable
xx$Urban_or_Rural_Area[xx$Urban_or_Rural_Area == "1"] <- "Urban"
xx$Urban_or_Rural_Area[xx$Urban_or_Rural_Area == "2"] <- "Rural"

# Remove Unallocated recods
xx = xx[!xx$Urban_or_Rural_Area == "3",]

# Draw a boxplot
ggplot(xx, aes(x=Urban_or_Rural_Area, y=Number_of_Vehicles))+geom_boxplot()

ggplot(xx, aes(x=Urban_or_Rural_Area, y=Number_of_Vehicles))+geom_boxplot()+scale_y_log10()
```

As it can be seen, in the urban areas...

A x test has been carried out to investigate whether the average number of vehicles in an accident differ in urban and rural areas.
 
```{r}
# Test to investigate whether the average number of vehicles differes per areas

```

As we can see, the...

## Statistical hypothesis

Using a suitable statistical hypothesis test, **investigate whether the frequency of accidents varies by day of the week**. Repeat this test using only week-days (excuding Saturday and Sunday).

In this section, we investigate whether the frequency of accidents varies by day of the week. To do this we use a statistical hypotesis test.

```{r}

```

This time the test is done using only week-days (excluding Saturday and Sunday).

```{r}

```

## Confidence interval

**Compute a 95% confidence interval** for the expected (mean) number of accidents which occur on a Monday. State your assumptions in computing this interval, and verify whether they are valid.

Finally, in this section, we compute a 95% confidence interval for the expected (mean) number of accidents which occur on a Monday.

```{r}

```

